Phenotypic evolution is more constrained in simpler food webs

### 2019-05-07 manuscript.Rmd
finished draft of revision and sent to Jordi for comments

\vspace*{0.1cm} 
\begin{center} \LARGE Extinction of a consumer guild constrains phenotypic evolution in an insect herbivore \end{center}

@@

Global change is altering the structure of ecological networks; however, we are currently in a poor position to predict how these altered communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that simulated the extinction of a guild of larval parasitoids that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that the number of traits under selection increased with the extinction of larval parasitoids. In contrast, the adaptive landscape was more neutral in the intact food web because different parasitoid guilds impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that the loss of trophic interactions can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. As a control (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. To exclude the guild of larval parasitoids, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989].
In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. Since we were interested in selection imposed by parasitoids, we excluded other sources of mortality (`r round((mean(full_df$early.larval.death) + mean(full_df$unknown_mortality)), 2)*100`% mortality from other sources). Another reason why we excluded other sources of mortality, such as early larval death, is that they can influence the expression of the gall phenotype and confound estimates of selection. For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

**Selection surface**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_{z_i}$), quadratic ($\alpha_{z_i:z_i}$), and linear interactions ($\alpha_{z_i:z_j}$) between each trait. We also allowed these trait-fitness relationships ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear trait-fitness relationships, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think that modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e., setting them to zero) when calculating trait-fitness relationships. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as trait-fitness relationships in each food-web treatment. To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.

@@

$$\textbf{C} = \begin{pmatrix} \gamma_{\text{Diam:Diam}}&& \\ \gamma_{\text{Clutch:Diam}}&\gamma_{\text{Clutch:Clutch}}& \\ \gamma_{\text{Pref:Diam}} & \gamma_{\text{Pref:Clutch}} &\gamma_{\text{Pref:Pref}} \end{pmatrix} - \begin{pmatrix} \beta_{\text{Diam}}\beta_{\text{Diam}}&& \\ \beta_{\text{Clutch}}\beta_{\text{Diam}}&\beta_{\text{Clutch}}\beta_{\text{Clutch}}& \\ \beta_{\text{Pref}}\beta_{\text{Diam}} & \beta_{\text{Pref}}\beta_{\text{Clutch}} &\beta_{\text{Pref}}\beta_{\text{Pref}} \end{pmatrix}$$
$$\textbf{C} = \begin{pmatrix} C_{\text{Diam:Diam}}&& \\ C_{\text{Clutch:Diam}} & C_{\text{Clutch:Clutch}} & \\ C_{\text{Pref:Diam}} & C_{\text{Pref:Clutch}} & C_{\text{Pref:Pref}} \end{pmatrix}$$

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. The sign of diagonal terms of the curvature matrix dictates whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (i.e., less independence). Therefore, we can infer the number of genetic constraints by counting the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix. For these analyses, we only retained selection gradients that showed clear evidence of being different from zero (i.e. 95% CI did not overlap zero) prior to calculating the curvature ($\gamma - \beta \beta^\text{T}$) in each food-web treatment. This provided us with a conservative estimate of the factors contributing to the curvature of the adaptive landscape and guarding against any false positive (i.e. Type I error).

Note that we do not put much stock in differences in the magnitude of selection between food-web treatments, but instead are concerned primarily with whether these gradients are clearly nonzero. This is because their is a high likelihood that we are underestimating the magnitude of each selection gradient in our treatment that excludes larval parasitoids. This occurs because larval parasitoids not only impose mortality on gall midges, but egg parasitoids as well (X% mortality). Therefore, our short-term removal of larval parasitoids has not allowed for a numerical response in egg parasitoids. An increase in the abundance of egg parasitoids would surely increase the magnitude of selection estimates in the extinction treatment due to the increased mortality of gall midges, which would increase variance in relative fitness. A recent paper by @Hunter2008 highlight this pervasive effect of survival on the magnitude of selection estimates. Therefore, we expect the magnitude of selection reflects a transient dynamic that would change over time.
This is why we put more stock in our qualitative estimates of selection (i.e. selective constraints = number of traits under directional selection; genetic constraints = number of negative variances + nonzero covariances).

@@

Rather than imposing selection, parasitoids may influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we modeled only a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (complex $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; simple $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's selection surface and directional selection gradient.

@@

Once parasitized, the gall phenotype becomes the phenotype of the egg parasitoid. This phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify selection surfaces and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate the selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Loss of trophic interactions increase selective constraints

\indent We found that more gall midge traits experienced directional selection in the simple (3 of 3) vs. complex food-web treatment (1 of 3)(table \ref{Table:Gradients}). For example, we observed directional selection for smaller clutch sizes in the simple food web, but there was no evidence of selection acting on this trait in the complex food web (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of selection appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids. Specifically, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes ($\beta_{\text{Clutch}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.high,2)`]). In the simple food web, we also observed clear evidence of directional selection for midges to avoid ovipositing on plants with high densities of conspecifics (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}B); however, there was no clear relationship in complex food webs (table \ref{Table:Gradients}). This was likely a result of larval parasitoids imposing greater mortality on egg parasitoids at high gall midge densities (see Selection on egg parasitoids section), and thus a less than additive effect on gall midges. Chamber diameter experienced positive directional selection in both food-web treatments, but selection was `r round(((filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean - filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) / filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) * 100, 0)`% higher in complex food webs (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}). This was not due to any difference between egg and larval parasitoids ($\beta_{\text{Diam}}$ contrast = 
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`]), but was due to the lower average survival in complex food webs ($\bar W$ contrast `r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$mean, 2)`
[`r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.low, 2)`,
`r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.high, 2)`]).

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in intact food webs vs. under consumer extinction.}
\label{Table:Gradients}
\centering
\begin{tabular}{lcc} %c
\\ 
\hline
\textbf{Selection gradient} & \textbf{Intact} & \textbf{Extinction}   \\ %& \textbf{Contrast}
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.high,2)`] }& 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.high,2)`] }\\%& 
%\textbf{
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`] }\\

@@

## Loss of trophic interactions increase genetic constraints

\indent The curvature of the adaptive landscape indirectly affects genetic constraints and is influenced by directional, quadratic, and correlational selection gradients. There was no clear evidence of correlational selection for any combination of traits in either food-web treatment (table \ref{Table:Gradients}). Similarly, there was no clear evidence of quadratic selection on chamber diameter or clutch size in either food-web treatment (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}A,C). In contrast, our food-web treatment did alter quadratic selection acting on oviposition preference (table \ref{Table:Gradients}). The negative relationship between oviposition preference and larval survival dampened at high densities in complex, but not simple food webs (fig. \ref{fig:UV_Landscape}B). This dampened relationship was partly due to a trend for nonlinear selection imposed by larval parasitoids ($\gamma_{\text{Pref:Pref}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$conf.high,2)`]), but was also magnified by the lower average survival in complex food webs.

```{r UV_Landscape, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:UV_Landscape}Adaptive landscape of gall midge phenotypes in intact food webs vs. consumer extinction. Each panel corresponds to a different phenotypic trait: chamber diameter (A); oviposition preference (B); and clutch size (C). Bold lines represent selection experienced in intact food webs (orange) and with consumer extinction (blue). Thin lines represent bootstrapped replicates to show the uncertainty in selection. For clarity, we only display 100 bootstraps even though inferences are based on 1,000 replicates. Note that mean larval survival is plotted on a natural log scale to reflect the mathematical defintion of the adaptive landscape."}
knitr::include_graphics("analyses/UV_landscapes.pdf")
```

Using our estimates of directional, quadratic, and correlation selection, we calculated the curvature ($\text{C}=\gamma - \beta \beta^\text{T}$) of the adaptive landscape in each food-web treatment. In order to be conservative in the factors contributing to curvature, we only used those estimates that clearly differed from zero (i.e. 95% CI did not overlap zero), otherwise we set the component gradients to zero. We found that the curvatures of the adaptive landscape exhibited the following structures:

@@

$$\textbf{C} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}$$

>>>>

\vspace*{0.1cm} 
\begin{center} \LARGE Consumer extinctions constrain phenotypic evolution in an insect herbivore \end{center}

@@

Global change is altering the structure of ecological networks; however, we are currently in a poor position to predict how these altered communities will affect the evolutionary potential of remaining populations. Theory on adaptive landscapes provides a framework for predicting how selection constrains phenotypic evolution, but often treats the community context of evolving populations as a ``black box''. Here, we integrate ecological networks and adaptive landscapes to examine how changes in food-web structure shape evolutionary constraints. We conducted a field experiment that simulated the extinction of a guild of larval parasitoids that were able to impose selection on an insect herbivore. We then measured herbivore survival as a function of three key phenotypic traits. We found that the number of traits under selection increased with the extinction of larval parasitoids. In contrast, the adaptive landscape was more neutral in the intact food web because different parasitoid guilds impose different selection pressures, minimizing relative fitness differences among phenotypes. Our results suggest that the loss of trophic interactions due to consumer extinctions can impose greater constraints on phenotypic evolution. This indicates that the simplification of ecological communities may constrain the adaptive potential of remaining populations to future environmental change.

@@

We setup our food-web manipulation across 128 plants soon after galls began developing on willows in early June of 2013. These 128 plants came from eight different plant genotypes that spanned the range of trait variation observed in this willow population [@Barbour2015]. For intact food webs (eight replicates per genotype), we used flagging tape to mark 14 galled leaves per plant (~30 larvae), allowing the full suite of egg and larval parasitoids to impose selection. Marking galls with flagging tape ensured that we compared control and treatment galls with similar phenology when we collected galls later in the season. To simulate consumer extinctions, we enclosed 14 galled leaves with 10x15cm organza bags (ULINE, Pleasant Prairie, WI, USA) to exclude three parasitoid species that attack during larval development. This treatment did not exclude the egg parasitoid *Platygaster* sp., which attacks prior to gall initiation [larva initiate gall development in Cecidomyiid midges: @Gagne1989].
In late August, we collected marked and bagged galls from each plant, placed them into 30 mL vials and kept them in the lab for 4 months at room temperature. We then opened galls under a dissecting scope and determined whether larvae survived to pupation (our measure of fitness) or were parasitized. We removed other sources of mortality, such as early larval death, as they could influence the expression of the gall phenotype and confound estimates of selection. <!--Since we were interested in selection imposed by parasitoids, we excluded other sources of mortality (`r round((mean(full_df$early.larval.death) + mean(full_df$unknown_mortality)), 2)*100`% mortality from other sources). Another reason why we excluded other sources of mortality, such as early larval death, is that they can influence the expression of the gall phenotype and confound estimates of selection.--> For the food-web treatment that excluded larval parasitoids, we further restricted our data by removing any incidental instances of parasitism by a larval parasitoid. This represented less than 3\% of the observations in this food-web treatment and allowed us to focus our inferences of selection on those imposed by the egg parasitoid. Our final dataset contains survival estimates for `r no_larvae` larvae from `r no_galls` galls and `r no_plants` plants.

@@

**Selection surfaces**: Since larval survival was our measure of absolute fitness, we used a GLMM that assumed a binomial error distribution (and logit-link function). To approximate the selection surface, we modelled larval survival as a function of food-web treatment as well as linear ($\alpha_{z_i}$), quadratic ($\alpha_{z_i:z_i}$), and linear interactions ($\alpha_{z_i:z_j}$) between each trait. We also allowed these selection surfaces ($\alpha$) to vary between food-web treatments. Note that to obtain valid estimates of linear selection surfaces, we removed nonlinear terms prior to estimating linear relationships [@Lande1983]. Other approaches have been advocated for approximating selection surfaces [@Schluter1988]; however, our approach enables us to calculate selection gradients, and thus is more appropriate for approximating the adaptive landscape [@Arnold2003]. To account for the nonindependence of clutch size (measured at gall level) and oviposition preference (measured at plant level) as well as any independent effects of willow genotype on larval survival, we modelled gall ID nested within plant ID nested within genotype ID as random effects. Although statistical models with random effects are not common in analyses of natural selection, we think that modelling random effects can mitigate biased estimates of selection due to environmental covariances between traits and fitness [@Rausher1992]. Since our end goal was to characterize the relationship between mean trait values and mean fitness (adaptive landscape), we assumed the mean value of our random effects (i.e., setting them to zero) when calculating selection surfaces. We then used parametric bootstrapping (1,000 replicates) to estimate the effect of food-web treatment on larval survival as well as selection surfaces in each food-web treatment. <!--To determine whether trait-fitness relationships differed between food-web treatments, we calculated the difference in bootstrapped replicates between treatments.-->

@@

$$\textbf{C} = \begin{pmatrix} \gamma_{\text{Diam:Diam}}&& \\ \gamma_{\text{Clutch:Diam}}&\gamma_{\text{Clutch:Clutch}}& \\ \gamma_{\text{Pref:Diam}} & \gamma_{\text{Pref:Clutch}} &\gamma_{\text{Pref:Pref}} \end{pmatrix} - \begin{pmatrix} \beta_{\text{Diam}}\beta_{\text{Diam}}&& \\ \beta_{\text{Clutch}}\beta_{\text{Diam}}&\beta_{\text{Clutch}}\beta_{\text{Clutch}}& \\ \beta_{\text{Pref}}\beta_{\text{Diam}} & \beta_{\text{Pref}}\beta_{\text{Clutch}} &\beta_{\text{Pref}}\beta_{\text{Pref}} \end{pmatrix}$$
$$\textbf{C} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}$$

Note that we ommitted the upper triangle of each matrix for clarity since it is simply the reflection of the lower triangle. The sign of diagonal terms of the curvature matrix dictates whether selection will increase (+), decrease (-), or cause no change (0) in the additive genetic variance of a trait. Similarly, any nonzero covariance terms (off-diagonal) are indicative of selection for trait integration (i.e., less independence). Therefore, we can infer the number of genetic constraints by summing the number of negative signs along the diagonal (decrease in additive genetic variance) and the number of nonzero terms along the off-diagonal (trait integration) of the curvature matrix. For these analyses, we retained estimates for selection gradients that showed clear evidence of being different from zero (i.e., 95% CI did not overlap zero) and set other gradients to zero, prior to calculating the curvature ($\gamma - \beta \beta^\text{T}$) in each food-web treatment. This provides a conservative estimate of the factors contributing to the curvature of the adaptive landscape.

Note that our metrics of selective and genetic constraints do not put much stock in differences in the magnitude of selection between food-web treatments. This is because there is a high likelihood that we are underestimating the magnitude of selection in our treatment that excludes larval parasitoids. We expect this because larval parasitoids not only impose mortality on gall midges, but egg parasitoids as well (fig. \ref{fig:Conceptual}). Therefore, our short-term removal of larval parasitoids has not allowed for an increase in egg parasitoid abundance after being released from intraguild predation. An increase in the abundance of egg parasitoids would decrease mean survival of gall midges, thus increasing the variance in relative fitness of midges, and consequently the magnitude of selection estimates. A recent paper by @Hunter2008 highlight this pervasive effect of survival on the magnitude of selection estimates.<!--Therefore, we expect the magnitude of selection reflects a transient dynamic that would change over time.--> This is why we use more qualitative metrics of evolutionary constraints (i.e., selective constraints = number of traits under directional selection; genetic constraints = number of decreasing variances + nonzero covariances).

@@

Rather than imposing selection, parasitoids may influence the expression of herbivore traits which could bias measurements of selection. In our system, it was plausible that parasitoids may influence chamber diameter by altering larval feeding behavior or killing larvae before they complete their development. To estimate this potential bias, we subset our data to only include galls where there was variation in larval survival within the same gall (i.e. 1 > mean survival > 0). If we assume that larvae within each gall should have similar chamber diameters because they come from the same clutch and experience the same local environment (an assumption our data supports: gall ID explains 54% of the variance in chamber diameter), then the relationship between chamber diameter and larval survival in this data subset represents the effect of parasitism on trait expression (i.e. bias). We used a GLMM with the same structure as described above except that we modeled only a linear relationship between chamber diameter and larval survival ($\alpha_{\text{Diam}}$). We detected a positive bias in both food-web treatments (intact $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebComplex:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebComplex:sc.Diam")$conf.high,2)`]; extinction $\alpha_{\text{Diam}}$=
`r round(fixef(biased_foodweb_model)["FoodwebSimple:sc.Diam"],2)` 
[`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.low,2)`,
`r round(filter(biased_foodweb_confint, term=="FoodwebSimple:sc.Diam")$conf.high,2)`]), indicating that unadjusted relationships would overestimate the strength of selection on chamber diameter. To account for this bias, we substracted our mean estimates of bias from our estimates with the full dataset prior to calculating chamber diameter's selection surface and directional selection gradient.

@@

Once parasitized, the gall phenotype becomes the phenotype of the egg parasitoid. This phenotype may influence the egg parasitoid's survival in the face of larval parasitoids, and thus experiences selection. Our food-web manipulation allows us to measure selection imposed by larval parasitoids on the phenotype of egg parasitoids. Using the same models as described above, we substituted egg parasitism as our response variable to quantify selection surfaces and selection gradients acting on the egg parasitoid. Note that we cannot test the effect of food-web structure on the egg parasitoid's adaptive landscape ---we can only estimate selection imposed by larval parasitoids. This comparison is still useful though in determining the extent to which the community context may have indirect evolutionary effects by altering selection on multiple interacting populations.

@@

## Consumer extinctions increase selective constraints

\indent We found that the extinction of larval parasitoids increased the number of gall midge traits under directional selection (3 of 3) relative to the intact food web (1 of 3)(table \ref{Table:Gradients}). For example, we observed directional selection for smaller clutch sizes when we excluded larval parasitoids, but there was no evidence of selection acting on this trait in the intact food web (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}C). This absence of selection appeared to be a result of conflicting selection pressures imposed by each guild of parasitoids. Specifically, when we subset our data to focus on differences between parasitoid guilds, we found that larval parasitoids actually impose directional selection for larger clutch sizes ($\beta_{\text{Clutch}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="sc.log.Clutch")$conf.high,2)`]). We also observed clear evidence of directional selection for midges to avoid ovipositing on plants with high densities of conspecifics when we excluded larval parasitoids (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}B); however, there was no clear relationship in the intact food web (table \ref{Table:Gradients}). This was likely a result of larval parasitoids imposing greater mortality on egg parasitoids at high gall midge densities (see Selection on egg parasitoids section), and thus a less than additive effect on gall midges. Chamber diameter experienced positive directional selection in both food-web treatments (fig. \ref{fig:UV_Landscape}A). Although the magnitude of selection on chamber diameter was relatively higher in the intact food web (table \ref{Table:Gradients}), this was not due to any difference in the relationship between chamber diameter and survival (selection surfaces: intact $\alpha_{\text{Diam}}$=
`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="sc.Diam")$mean,2)` 
[`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="sc.Diam")$conf.high,2)`]; extinction $\alpha_{\text{Diam}}$=
`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="sc.Diam")$mean,2)` 
[`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="sc.Diam")$conf.high,2)`]), but was simply a consequence of the (expected) lower survival in intact food webs (intact $\bar W$=
`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="(Intercept)")$mean,2)` 
[`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="(Intercept)")$conf.low,2)`,
`r round(filter(tidy_foodweb_alphas, type=="Complex", term=="(Intercept)")$conf.high,2)`]; extinction $\bar W$=
`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="(Intercept)")$mean,2)` 
[`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="(Intercept)")$conf.low,2)`,
`r round(filter(tidy_foodweb_alphas, type=="Simple", term=="(Intercept)")$conf.high,2)`]).<!-- but selection was `r round(((filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean - filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) / filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean) * 100, 0)`% higher in complex food webs (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}). This was not due to any difference between egg and larval parasitoids ($\beta_{\text{Diam}}$ contrast = 
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
`r round(-1*filter(tidy_egglarval_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`]), but was due to the lower average survival in complex food webs ($\bar W$ contrast `r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$mean, 2)`
[`r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.low, 2)`,
`r round(-1*filter(tidy_foodweb_alphas, type=="Diff", term=="(Intercept)")$conf.high, 2)`]).-->

@@

\begin{table}[h]
\caption{Standardized selection gradients acting on gall midges in the intact food web and with the extinction of larval parasitoids.}
\label{Table:Gradients}
\centering
\begin{tabular}{lcc} %c
\\ 
\hline
\textbf{Selection gradient} & \textbf{Intact} & \textbf{Extinction}   \\ %& \textbf{Contrast}
\hline
$\beta_{\text{Diam}}$ & 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Complex", term=="sc.Diam")$conf.high,2)`] }& 
\textbf{
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$mean,2)` [
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.low,2)`,
`r round(filter(tidy_foodweb_grads, type=="Simple", term=="sc.Diam")$conf.high,2)`] }\\%& 
%\textbf{
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$mean,2)` [
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.low,2)`,
%`r round(-1*filter(tidy_foodweb_grads, type=="Diff", term=="sc.Diam")$conf.high,2)`] }\\

@@

## Consumer extinctions increase genetic constraints

\indent The curvature of the adaptive landscape indirectly affects genetic constraints and is influenced by directional, quadratic, and correlational selection gradients. There was no clear evidence of correlational selection for any combination of traits in either food-web treatment (table \ref{Table:Gradients}). Similarly, there was no clear evidence of quadratic selection on chamber diameter or clutch size in either food-web treatment (table \ref{Table:Gradients}; fig. \ref{fig:UV_Landscape}A,C). In contrast, our food-web treatment did alter quadratic selection acting on oviposition preference (table \ref{Table:Gradients}). The negative relationship between oviposition preference and larval survival dampened at high densities in intact food webs, but not with the exclusion of larval parasitoids (fig. \ref{fig:UV_Landscape}B). This dampened relationship was partly due to a trend for nonlinear selection imposed by larval parasitoids ($\gamma_{\text{Pref:Pref}}$=
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$mean,2)` 
[`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$conf.low,2)`,
`r round(filter(tidy_egglarval_grads, type=="Complex", term=="I(sc.sqrt.Pref^2)")$conf.high,2)`]), but was also magnified by the lower average survival in complex food webs.

```{r UV_Landscape, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:UV_Landscape}Adaptive landscape of gall midge phenotypes in the intact food web and with the extinction of larval parasitoids. Each panel corresponds to a different phenotypic trait: chamber diameter (A); oviposition preference (B); and clutch size (C). Bold lines represent selection experienced in intact food webs (orange) and with consumer extinctions (blue). Thin lines represent bootstrapped replicates to show the uncertainty in selection. For clarity, we only display 100 bootstraps even though inferences are based on 1,000 replicates. Note that mean larval survival is plotted on a natural log scale to reflect the mathematical defintion of the adaptive landscape."}
knitr::include_graphics("analyses/UV_landscapes.pdf")
```

Using our estimates of directional, quadratic, and correlation selection, we calculated the curvature ($\text{C}=\gamma - \beta \beta^\text{T}$) of the adaptive landscape in each food-web treatment. In our study, this corresponds to:

@@

$$\textbf{C} = \begin{pmatrix} \text{C}_{\text{Diam:Diam}}&& \\ \text{C}_{\text{Clutch:Diam}} & \text{C}_{\text{Clutch:Clutch}} & \\ \text{C}_{\text{Pref:Diam}} & \text{C}_{\text{Pref:Clutch}} & \text{C}_{\text{Pref:Pref}} \end{pmatrix}$$
\bigskip

In order to obtain a conservative estimate of the factors contributing to the curvature, we only used estimates of selection that clearly differed from zero (bold values in table \ref{Table:Gradients}), otherwise we set estimates of selection to zero (non-bold values in table \ref{Table:Gradients}) prior to our calculation. We found that the curvatures of the adaptive landscape exhibited the following structures:
